---
title: "AML_Pipeline"
author: "Krish Shah"
date: "2023-10-20"
output:
  word_document: default
  html_document: default
  pdf_document: default
---
Import necessary packages and clear environment.
```{r}
rm(list=ls())
suppressPackageStartupMessages(require(tidymodels))
suppressPackageStartupMessages(require(tidyverse))
suppressPackageStartupMessages(require(workflows))
suppressPackageStartupMessages(require(tune))
suppressPackageStartupMessages(require(themis))
suppressPackageStartupMessages(require(vip))
suppressPackageStartupMessages(require(glue))
suppressPackageStartupMessages(require(eulerr))
suppressPackageStartupMessages(require(BiocParallel))
suppressPackageStartupMessages(require(RVenn))
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(ggpubr))
suppressPackageStartupMessages(require(ComplexHeatmap))
suppressPackageStartupMessages(require(circlize))
suppressPackageStartupMessages(require(RColorBrewer))
suppressPackageStartupMessages(require(DESeq2))
suppressPackageStartupMessages(require(GGally))
suppressPackageStartupMessages(require(ranger))
suppressPackageStartupMessages(require(logger))
suppressPackageStartupMessages(require(crayon))
suppressPackageStartupMessages(require(tibble))
suppressPackageStartupMessages(require(glue))
suppressPackageStartupMessages(require(doFuture))
suppressPackageStartupMessages(require(parallel))
suppressPackageStartupMessages(require(skimr))
suppressPackageStartupMessages(require(limma))
suppressPackageStartupMessages(require(preprocessCore))
suppressPackageStartupMessages(require(patchwork))
suppressPackageStartupMessages(require(xgboost))
suppressPackageStartupMessages(require(kernlab))
suppressPackageStartupMessages(require(caret))
```

Source machine learning functions from utils.R
```{r}
source("/Users/krishshah/Downloads/R markdown transfer files/utils.R")
```

Parallel computing for cores. Do not run on local R Studio.
```{r}
#options(future.globals.maxSize= 10 * 2^1024)
#n = Sys.getenv('LSB_MAX_NUM_PROCESSORS')
#all_cores <- as.numeric(n)

#registerDoFuture()
#cl <- makeCluster(all_cores)
#plan(future::cluster, workers = cl)
```

Load SJ + TARGET TPM data.
```{r}
TPM <- read.table("/Users/krishshah/Downloads/R markdown transfer files/SJAML842_TPM.txt",row.names = 1)
row_names = TPM$geneID
TPM = TPM[,-c(1:4)]
TPM = data.matrix(TPM)
rownames(TPM) = row_names
```

Load SJ + TARGET count data.
```{r}
count <- read.table("/Users/krishshah/Downloads/R markdown transfer files/SJAML842_count.txt")
counts_mat = data.matrix(count[,-1])
rownames(counts_mat) = count$geneID
rows = rownames(counts_mat) 
counts_mat = counts_mat[,-c(1:3)]
```

Load SJ + TARGET meta data.
```{r}
meta <- readr::read_tsv("/Users/krishshah/Downloads/R markdown transfer files/AML_n856_coreReferenceSetcases_update.txt", show_col_types = FALSE)
subset_meta <- subset(meta, Sample %in% colnames(TPM))
meta = subset_meta
```

Make sure meta lines up with count and TPM data + table on Source and Libstrand.
```{r}
all(colnames(TPM)==meta$Sample)
all(colnames(counts_mat)==meta$Sample)
table(meta$Source, meta$LibStrand)
```

Load AAML1031 count data.
```{r}
test1 = readr::read_tsv("/Users/krishshah/Downloads/R markdown transfer files/KLCO-AAML1031_n1034Dx-Batch01-STRANDED_RSEM_gene_count.2023-03-22_00-02-47.txt")
test2 = readr::read_tsv("/Users/krishshah/Downloads/R markdown transfer files/KLCO-AAML1031_n1034Dx-Batch02-STRANDED_RSEM_gene_count.2023-03-22_22-14-18.txt")
```

Combine both count files.
```{r}
AAML_data = cbind(test1, test2)
row_names = AAML_data$geneID
AAML_data = AAML_data[,-c(1:4)]
rownames(AAML_data) = row_names
```

Load in AAML1031 meta.
```{r}
AAML_meta = readxl::read_xlsx("/Users/krishshah/Downloads/R markdown transfer files/AAML1031_lcd5000_prediction.xlsx")
AAML_meta_simplified = data.frame(Sample = AAML_meta$ID, MolClass = AAML_meta$MolClass, LibStrand = "Stranded")
```

Load in AAML1031 TPM and match with meta file.
```{r}
TPM1 = read.table("/Users/krishshah/Downloads/R markdown transfer files/KLCO-AAML1031_n1034Dx-Batch01-STRANDED_RSEM_gene_TPM.2023-03-22_00-02-47.txt", row.names = 1)
TPM1 = TPM1[,-c(1:3)]
colnames(TPM1) = TPM1[1,]
TPM1 = TPM1[-1,]
TPM2 = read.table("/Users/krishshah/Downloads/R markdown transfer files/KLCO-AAML1031_n1034Dx-Batch02-STRANDED_RSEM_gene_TPM.2023-03-22_22-14-18.txt", row.names = 1)
TPM2 = TPM2[,-c(1:3)]
colnames(TPM2) = TPM2[1,]
TPM2 = TPM2[-1,]
a_TPM = cbind(TPM1,TPM2)
AAML_TPM = a_TPM[,AAML_meta$ID]
```

AAML1031 meta file has 1034 samples while count has 1038 so need to match up.
```{r}
AAML_data = AAML_data[,AAML_meta$ID]
all(colnames(AAML_data)==AAML_meta$ID)
all(colnames(AAML_TPM)==AAML_meta$ID)
```

Load AML17 TPM data.
```{r}
AML17_TPM <- read.table("/Users/krishshah/Downloads/R markdown transfer files/KLCO-AML_Remapped_hg38-STRANDED_RSEM_gene_TPM.2023-03-11_07-49-42.txt",row.names = 1)
AML17_TPM = AML17_TPM[-c(1:3)]
colnames(AML17_TPM) = AML17_TPM[1,]
AML17_TPM = AML17_TPM[-1,]

extra_TPM <- read.table("/Users/krishshah/Downloads/R markdown transfer files/SJAML031413_D1_TPM.txt")
extra_TPM = extra_TPM[,-c(2:4)] %>% as_tibble()
colnames(extra_TPM) = extra_TPM[1,]
extra_TPM = extra_TPM[-1,]
row_names = extra_TPM$geneID
extra_TPM = extra_TPM[,-1]
rownames(extra_TPM) = row_names

TPM_17 = cbind(AML17_TPM,extra_TPM)

row_names = colnames(TPM_17)
```

Only keep samples that are not in SJ+TARGET data or AAML1031 data.
```{r}
`%notin%` <- function(x, y) {
  !(x %in% y)
}

TPM_17 = t(TPM_17) %>% data.frame()
rownames(TPM_17) = row_names
TPM_17a = filter(TPM_17, rownames(TPM_17) %notin% meta$Sample)
TPM_17a = filter(TPM_17a, rownames(TPM_17a) %notin% AAML_meta$ID)
```

Load in AML17 count data.
```{r}
AML16_counts = read.table("/Users/krishshah/Downloads/R markdown transfer files/KLCO-AML_Remapped_hg38-STRANDED_RSEM_gene_count.2023-03-11_07-49-42.txt")
AML16_counts = AML16_counts[,-c(2:4)]
colnames(AML16_counts) = AML16_counts[1,]
AML16_counts = AML16_counts[-1,]
row_names = AML16_counts$geneID
AML16_counts = AML16_counts[,-1]
rownames(AML16_counts) = row_names
row_names = colnames(AML16_counts)
AML16_counts = t(AML16_counts) %>% as_tibble()
rownames(AML16_counts) = row_names
extra_count = read.table("/Users/krishshah/Downloads/R markdown transfer files/SJAML031413_D1.RSEM.genes.counts")
extra_count = extra_count[,-c(2:4)]
extra_count = extra_count[,-c(3:4)]
extra_count = t(extra_count) %>% as_tibble()
extra_count = extra_count[,-1]
colnames(extra_count) = extra_count[1,]
extra_count=extra_count[-1,]
rownames(extra_count) = "SJAML031413_D1"
AML16_countse = rbind(AML16_counts, extra_count)
```

Filter AML17 count data to not include data from SJ, TARGET, or AAML.
```{r}
AML_17c = filter(AML16_countse, rownames(AML16_countse) %notin% meta$Sample)
AML_17c = filter(AML_17c, rownames(AML_17c) %notin% AAML_meta$ID)
all(rownames(AML_17c)==rownames(TPM_17a))
```

Parse through and create SJ meta, SJ TPM, and SJ count.
```{r}
SJ_meta <- meta[meta$Source == "SJ",]
SJ_counts_mat = counts_mat[,SJ_meta$Sample]
SJ_TPM = TPM[,SJ_meta$Sample]
```

Parse through and create TARGET meta, TARGET TPM, and TPM count.
```{r}
TARGET_meta <- meta[meta$Source == "TARGET",]
TARGET_counts = counts_mat[, TARGET_meta$Sample]
TARGET_TPM = TPM[,TARGET_meta$Sample]
```

Check dimensions of count + meta data.
```{r}
dim(SJ_counts_mat)
dim(TARGET_counts)
dim(AAML_data)
dim(AML_17c)
dim(SJ_meta)
dim(TARGET_meta)
dim(AAML_meta)
```

Make sure each file is samples x genes and in tibble structure.
```{r}
SJ_counts = t(SJ_counts_mat) %>% as_tibble()
TARGET_counts = t(TARGET_counts) %>% as_tibble()
AAML_data = t(AAML_data) %>% as_tibble()
```

Create simplified meta files for each source data to merge for PCA.
```{r}
SJ_meta_simple = data.frame(Sample = SJ_meta$Sample,
                            Subtype = SJ_meta$MolClass,
                            LibraryPrep = SJ_meta$LibStrand,
                            Source = "SJ")
TARGET_meta_simple = data.frame(Sample = TARGET_meta$Sample,
                                Subtype = TARGET_meta$MolClass,
                                LibraryPrep = TARGET_meta$LibStrand,
                                Source = "TARGET")
AAML_meta_simple = data.frame(Sample = AAML_meta$ID,
                              Subtype = AAML_meta$MolClass,
                              LibraryPrep = "Stranded",
                              Source = "AAML1031")
AML17_meta = data.frame(Sample = rownames(AML_17c),
                        Subtype = "Unknown",
                        LibraryPrep = "Stranded",
                        Source = "AML17")
```

Merge count files and meta files and check dimensions.
```{r}
all_4_counts = rbind(SJ_counts, TARGET_counts, AAML_data, AML_17c)
all_4_meta = rbind(SJ_meta_simple, TARGET_meta_simple, AAML_meta_simple, AML17_meta)
dim(all_4_counts)
dim(all_4_meta)
```

Only take the 20,004 protein coding genes from the dataframe.
```{r}
geneInfo = readr::read_tsv("/Users/krishshah/Downloads/R markdown transfer files/MP2PRT_used_count.txt")[,1:4]

geneInfo_filtered <- geneInfo %>% 
  filter(geneID %in% names(all_4_counts))

# Filter columns of train_x based on geneInfo$type
all_4_counts_protein <- all_4_counts %>%
  select(geneInfo_filtered$geneID[geneInfo_filtered$bioType == "protein_coding"]) %>% data.matrix()
```

Create PCA df.
```{r}
pca_data = prcomp((log10(all_4_counts_protein+0.001)),retx = T,scale. = T,center = T)

percentVar <- pca_data$sdev^2 / sum(pca_data$sdev^2)

pca.df = cbind(pca_data$x[,1:15], all_4_meta) %>% as_tibble()
```

Plot by Source.
```{r}
p1 = ggplot(pca.df, aes(x = PC1,y=PC2, color = Source))+ 
  geom_point(size=1.5) + 
  geom_vline(xintercept = 50, linetype = "dashed", color = "red") +
  xlab(paste0("PC1: ",round(percentVar[1] * 100,digits = 2),"% variance")) +
  ylab(paste0("PC2: ",round(percentVar[2] * 100,digits = 2),"% variance")) +
  #ggrepel::geom_text_repel(force=4,size=3,show.legend = F) +
  theme_bw() + 
  ggsci::scale_color_d3() +
  ggtitle("Count-based PCA",subtitle = "20004 Protein Coding Genes") +
  theme(plot.title = element_text(hjust = 0.5,size = 10),
        plot.subtitle = element_text(hjust = 0.5,size = 8))

p1
```

Plot by Library prep protocol.
```{r}
p2 = ggplot(pca.df, aes(x = PC1,y=PC2, color = LibraryPrep))+ 
  geom_point(size=1.5) + 
  geom_vline(xintercept = 50, linetype = "dashed", color = "red") +
  xlab(paste0("PC1: ",round(percentVar[1] * 100,digits = 2),"% variance")) +
  ylab(paste0("PC2: ",round(percentVar[2] * 100,digits = 2),"% variance")) +
  #ggrepel::geom_text_repel(force=4,size=3,show.legend = F) +
  theme_bw() + 
  ggsci::scale_color_d3() +
  ggtitle("Count-based PCA",subtitle = "20004 Protein Coding Genes") +
  theme(plot.title = element_text(hjust = 0.5,size = 10),
        plot.subtitle = element_text(hjust = 0.5,size = 8))

p2
```

Check dimensions of TPM data so that we can do the same process.
```{r}
dim(TPM_17a)
dim(SJ_TPM)
dim(TARGET_TPM)
dim(AAML_TPM)
```

Combine TPM data in same order as meta file above.
```{r}
TPM_17 = t(TPM_17a) %>% as_tibble()
all_4_TPM = t(cbind(SJ_TPM, TARGET_TPM, AAML_TPM, TPM_17)) %>% as_tibble()
```

Only choose protein-coding genes.
```{r}
geneInfo = readr::read_tsv("/Users/krishshah/Downloads/R markdown transfer files/MP2PRT_used_count.txt")[,1:4]

geneInfo_filtered <- geneInfo %>% 
  filter(geneID %in% names(all_4_TPM))

# Filter columns of train_x based on geneInfo$type
all_4_TPM_protein <- all_4_TPM %>%
  select(geneInfo_filtered$geneID[geneInfo_filtered$bioType == "protein_coding"]) %>% data.matrix()
```

Create PCA df using TPM input.
```{r}
pca_data = prcomp((log10(all_4_TPM_protein+0.001)),retx = T,scale. = T,center = T)

percentVar <- pca_data$sdev^2 / sum(pca_data$sdev^2)

pca.df = cbind(pca_data$x[,1:15], all_4_meta) %>% as_tibble()
```

Plot by Source.
```{r}
p1 = ggplot(pca.df, aes(x = PC1,y=PC2, color = Source))+ 
  geom_point(size=1.5) + 
  xlab(paste0("PC1: ",round(percentVar[1] * 100,digits = 2),"% variance")) +
  ylab(paste0("PC2: ",round(percentVar[2] * 100,digits = 2),"% variance")) +
  #ggrepel::geom_text_repel(force=4,size=3,show.legend = F) +
  theme_bw() + 
  ggsci::scale_color_d3() +
  ggtitle("TPM-based PCA",subtitle = "20004 Protein Coding Genes") +
  theme(plot.title = element_text(hjust = 0.5,size = 10),
        plot.subtitle = element_text(hjust = 0.5,size = 8))

p1
```

Plot by Library prep protocol.
```{r}
p2 = ggplot(pca.df, aes(x = PC1,y=PC2, color = LibraryPrep))+ 
  geom_point(size=1.5) + 
  xlab(paste0("PC1: ",round(percentVar[1] * 100,digits = 2),"% variance")) +
  ylab(paste0("PC2: ",round(percentVar[2] * 100,digits = 2),"% variance")) +
  #ggrepel::geom_text_repel(force=4,size=3,show.legend = F) +
  theme_bw() + 
  ggsci::scale_color_d3() +
  ggtitle("TPM-based PCA",subtitle = "20004 Protein Coding Genes") +
  theme(plot.title = element_text(hjust = 0.5,size = 10),
        plot.subtitle = element_text(hjust = 0.5,size = 8))

p2
```
